---
title: "Exercise 4.2 - Method of Moments"
format: pdf
editor: visual
---

## Exercise 4.2

The proportion of impurities in each manufactured unit of a certain kind of chemical product is a rv with pdf

$$
f(x)= 
\begin{cases}
    (\theta + 1)x^{\theta},& \text{if } 0 < x < 1 \\
    0,              & \text{otherwise}
\end{cases}
$$

where $\theta \geq -1$.

Five units of the manufactured product are taken in one day, resulting in the next impurity proportions: 0.33, 0.51, 0.02, 0.15, 0.12.

1.  Obtain the moment estimator of $\theta$.

## Solution

We start from the definition of the moment estimator, which we borrow from the main text.

------------------------------------------------------------------------

**Definition 4.1 (Method of moments)** Let $X∼F_θ$ with $θ=(θ_1,...,θ_K)′$. From a srs of X, the *method of moments* produces the estimator $\hat{θ}^{MM}$ that is the solution to the system of equations

$α_r(θ_1,…,θ_K)=a_r, \, r=1,…,R$

where $R \geq K$ is the lowest integer such that the system admits a unique solution and $θ_1,...,θ_K$ are the variables. The estimator $\hat{θ}^{MM}$ is simply referred to as the *moment estimator* of $\theta$.

------------------------------------------------------------------------

We recognise that $θ$ is just one parameter, so $R \geq 1$. To find the moment estimator, we start with one equation (R = 1) and check if that is enough. Therefore, we have the single equation:

$$
\alpha_1(\hat{θ}^{MM}) = a_1
$$ {#eq-method-moments}

with $\alpha_1$ and $a_1$ the population and sample mean, respectively. What we aim to do is to 'plug out' $\hat{θ}^{MM}$ from the expression above. We calculate the population moment first:

$$
\alpha_1 \equiv E(X) = \int_0^{1} xf(x)
=  \int_0^{1} (\theta + 1)x^{\theta + 1} dx = \frac{\theta + 1}{\theta + 2} |x^{\theta + 2}|_0^1 = \frac{\theta + 1}{\theta + 2}
$$ {#eq-first-moment}

Combining equations @eq-method-moments and @eq-first-moment we can express $\theta$ in terms of $a_1$: $$
\frac{\hat{θ}^{MM} + 1}{\hat{θ}^{MM} + 2} = a_1 \Longleftrightarrow  \hat{θ}^{MM} = \frac{2a_1 - 1}{1 - a_1}
$$ {#eq-estimator}

This shows that only one equation ($R = 1$) should be enough. We can now use `R` to compute the sample mean $a_1$ and plug it into equation @eq-estimator to find:

```{r}
impurity_props <- c(0.33, 0.51, 0.02, 0.15, 0.12)
a_1 <- mean(impurity_props)
print(paste("a_1 equals", a_1))
theta_mm <- (2*a_1 - 1)/(1 - a_1)
theta_mm
```

We find that that $a_1 = 0.226$ and $\hat{θ}^{MM} \approx -.71$.

### Visual Check

Let us try to reproduce our own version of Figure 4.2 in the main text. The strategy is as follows:

1.  For different $\theta$s, plot the density functions with their means as striped vertical lines
2.  Plot the observed sample mean, $a_1$ as a solid vertical line.
3.  Check visually whether the mean of the distribution of the selected $\hat{θ}^{MM}$ (striped black line) matches the solid black vertical line representing the observed sample mean.

```{r, fig.height=5}
### PREP
x_axis <- seq(0.1, 1, length.out = 1000)
thetas <- c(-.8, 0, 1)

# Function to compute density
f <- function(x, theta) {
  return(
    (theta + 1) * x^theta
    )
}
# Function to compute first pop moment (alpha_1) from theta (Eq. 3)
alpha_1 <- function(theta) {
  return(
    (theta + 1) / (theta + 2)
  )
}

### PLOTTING
# First Step: plot the mm estimate
plot(x = x_axis, y = f(x_axis, theta_mm), type = "l", xlab = "x", ylab = "Density")
abline(v = alpha_1(theta_mm), lty = 2)

# Second Step: Display a_1 as a vertical line
abline(v = a_1)
mtext(expression(a[1]), side = 1, at = a_1)

# Third Step: Loop over other thetas, overlay their density and first moment
i <- 1
rainbow_colours <- rainbow(length(thetas))
for (theta in thetas) {
  lines(x = x_axis, y = f(x_axis, theta), col = rainbow_colours[i])
  abline(v = alpha_1(theta), col = rainbow_colours[i], lty = 2)
  i <- i + 1
}

# Final Step: add legend
legend(
  x = "topright",
  legend = c(expression(hat(theta)[mm] == -.71), round(thetas, 2)),
  col = c("#000000", rainbow_colours),
  lwd = 2,
)
```

Since the two black lines (striped for population and solid for sample) overlay, we should have chosen the right moment estimator $\hat{θ}^{MM} \approx 0.71$.
